---
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
# This sets the overall chunk and output style
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README_files/",
  out.width = "100%"
)

# This hook enables reference to variables created later on (e.g. in abstract)
# To refer to such variables, use <%= variable %> or <% code %> (?brew::brew)
knitr::knit_hooks$set(document = function(x){
  x1 = paste(x, collapse = '\n')
  paste(capture.output(brew::brew(text = x1)), collapse = '\n')
})
```

# Quick sorting with R

**Author**: Xiurui Zhu<br />
**Modified**: `r file.info("README.Rmd")$mtime`<br />
**Compiled**: `r Sys.time()`

## Introduction

Quick sorting includes a series of techniques that quickly sorts a numeric vector. In this paper, two common quick-sorting techniques were introduced: pivot sorting and heapsort.

## Question

Given a numeric vector `input_vec`, sort it by increasing order.

```{r question}
set.seed(999L)
input_vec <- sample.int(1000L, size = 100L)
input_vec
```

## Solutions

```{r libraries, message=FALSE}
library(tidyverse)
```

### Common function

Sorting involves a basic option of swapping numbers. We first defined a helper function with swapping and a helper function with result printing.

```{r swap-function}
swap <- function(vec, i, j) {
  stopifnot(
    is.integer(i) == TRUE,
    is.integer(j) == TRUE,
    dplyr::between(i, 1L, length(vec)) == TRUE,
    dplyr::between(j, 1L, length(vec)) == TRUE
  )
  if (i != j) {
    tmp <- vec[i]
    vec[i] <- vec[j]
    vec[j] <- tmp
  }
  vec
}

# Test
swap(1L:5L, 2L, 5L)
```

### Pivot sorting

Pivot sorting selects a pivot in the original vector and all the other elements are compared with the pivot. Then the sequence is ordered increasingly or decreasingly around the pivot. Recursively call the function on both sides of the sequence until there is at most 1 element on each side.

```{r pivot-sort}
pivot_sort <- function(vec, decreasing = FALSE, ...) {
  stopifnot(anyDuplicated(vec) == 0, length(vec) > 0)
  arg_list <- list(...)
  # Initialize operation count
  if ("opn" %in% names(arg_list) == TRUE) {
    opn <- arg_list[["opn"]]
  } else {
    opn <- 0L
  }
  # Take first element as pivot
  pivot <- vec[1L]
  # Compare the rest with pivot
  larger_lgl <- vec[-1L] > pivot
  # Add operation count
  opn <- opn + length(vec)
  # Get elements on both sides
  left_vec <- vec[-1L][larger_lgl == decreasing]
  right_vec <- vec[-1L][larger_lgl != decreasing]
  # Recursively order elements on both sides
  if (length(left_vec) > 1L) {
    left_res <- pivot_sort(left_vec, decreasing, opn = opn)
    left_vec <- left_res[["vec"]]
    opn <- left_res[["opn"]]
  }
  if (length(right_vec) > 1L) {
    right_res <- pivot_sort(right_vec, decreasing, opn = opn)
    right_vec <- right_res[["vec"]]
    opn <- right_res[["opn"]]
  }
  # Output ordered vector
  list(vec = c(left_vec, pivot, right_vec), opn = opn)
}

# Test
pivot_sort(input_vec, decreasing = FALSE)
```

### Heapsort

Heapsort builds a binary tree of *n* layers out of the vector. At first, it tries to keep the root larger than the leaves (when sorting increasingly) by interating the comparison among root and leaves for *n* times, starting from layer *i* for the *i*th iteration. After traversing all roots, the top root is removed as the largest element in the vector and the last element is used to take its place, keeping the other nodes/leaves unchanged. Then, the new top root goes down through the tree if it is smaller than any of its leaves until it cannot move any more. After that, the top root is removed again and the last element is used to take its place. Repeat the process recursively until the length of the vector is shortened to 1.

```{r heapsort}
# Main function
heap_sort <- function(vec, decreasing = FALSE, ...) {
  stopifnot(anyDuplicated(vec) == 0, length(vec) > 0)
  arg_list <- list(...)
  # Initialize tail length
  if ("tail" %in% names(arg_list) == TRUE) {
    tail <- arg_list[["tail"]]
  } else {
    tail <- 0L
  }
  # Initialize operation count
  if ("opn" %in% names(arg_list) == TRUE) {
    opn <- arg_list[["opn"]]
  } else {
    opn <- 0L
  }
  # Split vector to sort
  vec_sort <- vec[1L:(length(vec) - tail)]
  if (tail > 0L) {
    vec_tail <- vec[(length(vec) - tail + 1L):length(vec)]
  } else {
    vec_tail <- NULL
  }
  if (length(vec_sort) > 1L) {
    if (tail == 0L) {
      # Initial setup by comparing root with leaves from bottom to top
      # After recur_idx iterations, the recur_idx-th layer is sorted
      order_res <- purrr::reduce(
        1:(ceiling(log2(length(vec_sort))) - 1L),
        function(order_res, recur_idx) {
          floor(length(vec_sort) / 2L):(2^(recur_idx - 1L)) %>%
            purrr::reduce(
              ~ {
                vec_tmp <- .x[["vec"]]
                opn <- .x[["opn"]]
                vec_tmp <- swap_heap(vec_tmp, decreasing, .y)
                vec_tmp <- purrr::pluck(vec_tmp, "vec")
                list(vec = vec_tmp, opn = opn + 1L)
              },
              .init = order_res
            )
        },
        .init = list(vec = vec_sort, opn = opn)
      )
      vec_sort <- order_res[["vec"]]
      opn <- order_res[["opn"]]
    } else {
      # Go through the heap in a top-down manner
      cur_idx <- 1L
      while(2L * cur_idx <= length(vec_sort)) {
        vec_sort <- swap_heap(vec_sort, decreasing, cur_idx)
        # Update current idx if swap has happened
        vec_status <- purrr::pluck(vec_sort, "status")
        vec_sort <- purrr::pluck(vec_sort, "vec")
        opn <- opn + 1L
        if (vec_status == 0L) {
          break
        } else {
          cur_idx <- cur_idx * 2L - 1L + vec_status
        }
      }
    }
    # Swap the first and the last elements of vec_sort
    vec_new <- swap(vec_sort, 1L, length(vec_sort))
    # Add sorted tail
    vec_new <- c(vec_new, vec_tail)
    # Here, opn + 1L stands for the operation of swapping in vec_new
    heap_sort(vec_new, decreasing, tail = tail + 1L, opn = opn + 1L)
  } else {
    # Stop recursion
    list(vec = vec, opn = opn)
  }
}
# Unit heap swap function
swap_heap <- function(vec, decreasing = FALSE, cur_idx) {
  status <- 0L
  if (decreasing == TRUE) {
    comp_idx <- which.min(vec[
      (2L * cur_idx):min(2L * cur_idx + 1L, length(vec))
    ])
    sel_idx <- comp_idx + 2L * cur_idx - 1L
    if (vec[cur_idx] > vec[sel_idx]) {
      vec <- swap(vec, cur_idx, sel_idx)
      status <- comp_idx
    }
  } else {
    comp_idx <- which.max(vec[
      (2L * cur_idx):min(2L * cur_idx + 1L, length(vec))
    ])
    sel_idx <- comp_idx + 2L * cur_idx - 1L
    if (vec[cur_idx] < vec[sel_idx]) {
      vec <- swap(vec, cur_idx, sel_idx)
      status <- comp_idx
    }
  }
  list(vec = vec, status = status)
}

# Test
heap_sort(input_vec, decreasing = FALSE)
```

